---
title: 'Project1 : Forecasting'
author: Jered Ataky, Matthew Baker, Christopher Bloome, David Blumenstiel, Dhairav
  Chhatbar
date: "6/6/2021"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Libraries
```{r}
library(dplyr)
library(forecast)
```

### Your data is a de-identified Excel spreadsheet.  Your assignment is to perform the appropriate analysis to forecast several series for 140 periods.  You will have 1622 periods for your analysis.

```{r}
df <- read.csv("https://raw.githubusercontent.com/davidblumenstiel/data/master/Data614Project1Dataset.csv")
colnames(df)[1] <- "SeriesIndex"
summary(df)

```

```{r}
df %>% 
  group_by(group) %>%
  summarise(var1_NAsum = sum(is.na(Var01)),
            var2_NAsum = sum(is.na(Var02)),
            var3_NAsum = sum(is.na(Var03)),
            var5_NAsum = sum(is.na(Var05)),
            var7_NAsum = sum(is.na(Var07)))
```

Missing data is distributed within the same observations for the most part.  Group has negligable effect on missing values.  Coincidentally, there are almost 140 missing observations per group; approximately the number we are supposed to forecast.
 
Seperating data by group and making timeseries

```{r}
#There's probably a prettier way to do this 
s01 <- subset(df[,c("group","Var01","Var02")], group == "S01")
s01$group <- NULL 
s01 <- ts(s01) 

s02 <- subset(df[,c("group","Var02","Var03")], group == "S02")
s02$group <- NULL 
s02 <- ts(s02) 

s03 <- subset(df[,c("group","Var05","Var07")], group == "S03")
s03$group <- NULL 
s03 <- ts(s03) 

s04 <- subset(df[,c("group","Var01","Var02")], group == "S04")
s04$group <- NULL 
s04 <- ts(s04) 

s05 <- subset(df[,c("group","Var02","Var03")], group == "S05")
s05$group <- NULL 
s05 <- ts(s05) 

s06 <- subset(df[,c("group","Var05","Var07")], group == "S06")
s06$group <- NULL 
s06 <- ts(s06) 

series <- list(s01, s02, s03, s04, s05, s06)

```

```{r}
#Could do this cleaner, but oh well
autoplot(series[[1]][,1], main =  "s01, var 1")
autoplot(series[[1]][,2], main =  "s01, var 2")
autoplot(series[[2]][,1], main =  "s02, var 2")
autoplot(series[[2]][,2], main =  "s02, var 3")
autoplot(series[[3]][,1], main =  "s03, var 5")
autoplot(series[[3]][,2], main =  "s03, var 7")
autoplot(series[[4]][,1], main =  "s04, var 1")
autoplot(series[[4]][,2], main =  "s04, var 2")
autoplot(series[[5]][,1], main =  "s05, var 2")
autoplot(series[[5]][,2], main =  "s05, var 3")
autoplot(series[[6]][,1], main =  "s06, var 5")
autoplot(series[[6]][,2], main =  "s06, var 7")

```
