---
title: "Data 624: Predictive Analytics - Assignment 1"
author: "Dhairav Chhatbar"
output:
  html_document: default
  word_document: default
---

```{r, message=FALSE, warning=FALSE}
library(fpp2)
library(readxl)
library(httr)
library(seasonal)
```


### 2.1

Use the help function to explore what the series gold, woolyrnq and gas represent.

**a) Use autoplot() to plot each of these in separate plots.**  

Daily morning gold prices in US dollars. 1 January 1985 – 31 March 1989
```{r}
autoplot(gold)
```
Quarterly production of woollen yarn in Australia: tonnes. Mar 1965 – Sep 1994.
```{r}
autoplot(woolyrnq)
```

Australian monthly gas production: 1956–1995.
```{r}
autoplot(gas)
```


b) What is the frequency of each series? Hint: apply the frequency() function.
```{r}
frequency(gold)
frequency(woolyrnq)
frequency(gas)
```


c) Use which.max() to spot the outlier in the gold series. Which observation was it?
```{r}
which.max(gold)
which.max(woolyrnq)
which.max(gas)
```

### 2.3
  
Download some monthly Australian retail data from the book website. These represent retail sales in various categories for different Australian states, and are stored in a MS-Excel file.
```{r, message=FALSE, warning=FALSE}
url<-"https://otexts.com/fpp2/extrafiles/retail.xlsx"
temp.file <- paste(tempfile(),".xlsx",sep = "")
download.file(url, temp.file, mode = "wb")

retaildata <- read_excel(temp.file, skip = 1)
```

b) Select one of the time series as follows (but replace the column name with your own chosen column)
```{r}
myts <- ts(retaildata[,"A3349873A"], frequency=12, start=c(1982,4))
```

**c) Explore your chosen retail time series using the following functions: autoplot(), ggseasonplot(), ggsubseriesplot(), gglagplot(), ggAcf(). Can you spot any seasonality, cyclicity and trend? What do you learn about the series?**  
There appears to be seasonality in the data as can be seen by the periodic spike in sales, and generally an increasing trend which dips a bit around 2000 and 2010. November and December are the highest sales points on the years. 
```{r}
autoplot(myts)
ggseasonplot(myts)
ggsubseriesplot(myts)
gglagplot(myts)
ggAcf(myts)
```





### 6.2

The plastics data set consists of the monthly sales (in thousands) of product A for a plastics manufacturer for five years.
```{r}
plastics
```


**a) Plot the time series of sales of product A. Can you identify seasonal fluctuations and/or a trend-cycle?**
From the below plot we can see that spike around mid year and return to a low point at the end of the year where the seasonality beings again for the next year. The trend seems to be additive and increasing. 
```{r}
autoplot(plastics)
```

**b) Use a classical multiplicative decomposition to calculate the trend-cycle and seasonal indices.**
```{r}
pls <- plastics %>% decompose(type = "multiplicative")
autoplot(pls) + labs(x="Year", y="Sales", title="Decomposition of Plastic Sales")

```


**c) Do the results support the graphical interpretation from part a?**  
The trend and seasonality do hold from part a. 

**d) Compute and plot the seasonally adjusted data.**
```{r}
autoplot(pls, series="Data") + autolayer(seasadj(pls), series="Seasonally Adjusted")
```

**e) Change one observation to be an outlier (e.g., add 500 to one observation), and recompute the seasonally adjusted data. What is the effect of the outlier?**  
The outlier causes a spike in the time series, but not the overall trend. If we adjust the for seasonality then the spike can be seen in the trend. 
```{r}
plastics2 <- plastics
plastics2[30] <- plastics2[30] + 500
pls2 <- plastics2 %>% decompose(type = "multiplicative")
autoplot(pls2, series="Data") + autolayer(seasadj(pls2), series="Seasonally Adjusted")

```

**f) Does it make any difference if the outlier is near the end rather than in the middle of the time series?**  
There is no difference between the the spike added in the middle of the series or the end, the behavior is still the same. 
```{r}
plastics3 <- plastics
plastics3[55] <- plastics3[55] + 500
pls3 <- plastics3 %>% decompose(type = "multiplicative")
autoplot(pls3, series="Data") + autolayer(seasadj(pls3), series="Seasonally Adjusted")

```

