---
title: "Predictive Analysis: First Half Semester Homework"
author: "Jered Ataky, Matthew Baker, Christopher Bloome, David Blumenstiel, Dhairav Chhatbar"
date: "6/6/2021"  
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Libraries

```{r warning=FALSE}

library(fpp2)

```


## Problem 1: HA 2.1

Use the help function to explore what the series gold, woolyrnq and gas represent.

**Description of each of the series:**

gold: Daily morning gold prices in US dollars. 1 January 1985 – 31 March 1989.
woolyrnq: Quarterly production of woollen yarn in Australia: tonnes. Mar 1965 – Sep 1994.
gas: Australian monthly gas production: 1956–1995.


a. Use autoplot() to plot each of these in separate plots.

**Plotting gold**

```{r}

autoplot(gold)

```


**Plotting woolyrnq**

```{r}

autoplot(woolyrnq)

```


**Plotting gas**

```{r}

autoplot(gas)

```



b. What is the frequency of each series? Hint: apply the frequency() function.

```{r}

frequency(gold)

```

gold data are annual.


```{r}

frequency(woolyrnq)

```
woolyrnq data are quarterly


```{r}

frequency(gas)

```

gas data are monthly. 


c. Use which.max() to spot the outlier in the gold series. Which observation was it?


```{r}

g <- which.max(gold)
g

```
gold get the max value at 770

Calculating gold value at t = 770:


```{r}

gold[g]
```

gold maximum value is 593.7


## Problem 2: HA 2.3

Download some monthly Australian retail data from the book website. These represent retail sales in various categories for different Australian states, and are stored in a MS-Excel file.

a. You can read the data into R with the following script:

```{r}

retaildata <- readxl::read_excel("retail.xlsx", skip=1)

```

Check the head of the retail data:

```{r}

head(retaildata)

```

The second argument (skip=1) is required because the Excel sheet has two header rows.


b. Select one of the time series as follows (but replace the column name with your own chosen column):

WE are going to use the column "A3349398A"

```{r}

myts <- ts(retaildata[,"A3349398A"],
  frequency=12, start=c(1982,4))


```



c. Explore your chosen retail time series using the following functions:

autoplot(), ggseasonplot(), ggsubseriesplot(), gglagplot(), ggAcf()

Can you spot any seasonality, cyclicity and trend? What do you learn about the series?


**Exploring A3349398A:"**

Plotting myts autoplot

```{r}
autoplot(myts)

```


Plotting myts ggseasonplot

```{r}
ggseasonplot(myts)

```



Plotting myts ggsubseriesplot

```{r}
ggsubseriesplot(myts)

```


Plotting myts gglagplot

```{r}
gglagplot(myts)

```

Plotting mts ggAcf

```{r}
ggAcf(myts)

```


We can see that there is a long-term increase and the time series is affected by 
a seasonal factor. Thus we observe both the **trend** and **seasonality**


## Problem 3: HA 6.2


The plastics data set consists of the monthly sales (in thousands) of product A for a plastics manufacturer for five years.

a. Plot the time series of sales of product A. Can you identify seasonal fluctuations and/or a trend-cycle?

b. Use a classical multiplicative decomposition to calculate the trend-cycle and seasonal indices.

c. Do the results support the graphical interpretation from part a?

d. Compute and plot the seasonally adjusted data.

e. Change one observation to be an outlier (e.g., add 500 to one observation), and recompute the seasonally adjusted data. What is the effect of the outlier?

f. Does it make any difference if the outlier is near the end rather than in the middle of the time series?






## Problem 4







## Problem 5








## Problem 6







## Problem 7






## Problem 8







## Problem 9







## Problem 10








## Problem 11

