---
title: "HW2"
author: "Christopher Bloome"
date: "7/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 6.3
**A chemical manufacturing process for a pharmaceutical product was discussed in Sect. 1.4. In this problem, the objective is to understand the relationship between biological measurements of the raw materials (predictors), measurements of the manufacturing process (predictors), and the response of product yield. Biological predictors cannot be changed but can be used to assess the quality of the raw material before processing. On the other hand, manufacturing process predictors can be changed in the manufacturing process. Improving product yield by 1 % will boost revenue by approximately one hundred thousand dollars per batch:**

## (a) 
**Start R and use these commands to load the data:**
```{r}
library(AppliedPredictiveModeling)
data(ChemicalManufacturingProcess)
```

**The matrix processPredictors contains the 57 predictors (12 describing the input biological material and 45 describing the process predictors) for the 176 manufacturing runs. yield contains the percent yield for each run.**

## (b) 

**A small percentage of cells in the predictor set contain missing values. Use an imputation function to fill in these missing values (e.g., see Sect. 3.8).** 

```{r}
library(caret)
library(RANN)
library(tidyverse)

prep <-preProcess(ChemicalManufacturingProcess[,-1], method = "knnImpute",
                  k = 5, knnSummary = mean)

CMP <- predict(prep, ChemicalManufacturingProcess, na.action = na.pass)

summary(CMP)

```

## (c) 
 
**Split the data into a training and a test set, pre-process the data, and tune a model of your choice from this chapter.** 

```{r}

set.seed(321)


trainR <- createDataPartition(CMP$Yield, p = .80, list = FALSE)
CMP.train <- CMP[trainR,]
CMP.test <- CMP[-trainR,]
```

```{r}
YieldX <-
  which(colnames(CMP) == "Yield")

controlx <- trainControl(method = "cv", number = 5)
modelx <- train(
  x = CMP.train[, -YieldX],
  y = CMP.train$Yield,
  method = "pls",
  tuneLength = 20,
  trControl = controlx
)
modelx
```

**What is the optimal value of the performance metric?**
```{r}
modelx[["bestTune"]][["ncomp"]]
```


## (d) 

**Predict the response for the test set. What is the value of the performance metric and how does this compare with the resampled performance metric on the training set?**

```{r}
postResample(predict(modelx, newdata = CMP.test, ncomp = mdl.ncomp), CMP.test$Yield)


getTrainPerf(modelx)


```

The RMSE Was actually better on the testing set. 

## (e) 

**Which predictors are most important in the model you have trained? Do either the biological or process predictors dominate the list?**

Manufacturing processes dominated the first 6 spots, ahead of a number of biological. 

```{r}
varImp(modelx)

```

## (f) 

**Explore the relationships between each of the top predictors and the response. How could this information be helpful in improving yield in future runs of the manufacturing process?**

We see that there is a relatively strong positive correlation in Manufacturing Process 32 and 09, while the other 3 in the top 5 have a negative correlation to Yield. Knowing the basics of how our processes contribute to Yield will help when diverting resources and making substitutions if ever needed. 

```{r}
CMP %>%
  ggplot(aes(x=ManufacturingProcess32, y=Yield)) +
  geom_point() + geom_smooth(method = 'lm')

```

```{r}
CMP %>%
  ggplot(aes(x=ManufacturingProcess36, y=Yield)) +
  geom_point() + geom_smooth(method = 'lm')

```
```{r}
CMP %>%
  ggplot(aes(x=ManufacturingProcess09, y=Yield)) +
  geom_point() + geom_smooth(method = 'lm')

```
```{r}
CMP %>%
  ggplot(aes(x=ManufacturingProcess13, y=Yield)) +
  geom_point() + geom_smooth(method = 'lm')

```
```{r}
CMP %>%
  ggplot(aes(x=ManufacturingProcess17, y=Yield)) +
  geom_point() + geom_smooth(method = 'lm')

```

# 7.2 

**Friedman (1991) introduced several benchmark data sets create by simulation. One of these simulations used the following nonlinear equation to create data: y = 10 sin(πx1x2) + 20(x3 − 0.5)2 + 10x4 + 5x5 + N(0, σ2) where the x values are random variables uniformly distributed between [0, 1] (there are also 5 other non-informative variables also created in the simulation). The package mlbench contains a function called mlbench.friedman1 that simulates these data:**

```{r}
library(mlbench)
set.seed(200)
trainingData <- mlbench.friedman1(200, sd = 1)
## We convert the 'x' data from a matrix to a data frame
## One reason is that this will give the columns names.
trainingData$x <- data.frame(trainingData$x)
## Look at the data using
featurePlot(trainingData$x, trainingData$y)
## or other methods.

## This creates a list with a vector 'y' and a matrix
## of predictors 'x'. Also simulate a large test set to
## estimate the true error rate with good precision:
testData <- mlbench.friedman1(5000, sd = 1)
testData$x <- data.frame(testData$x)

```

**Tune several models on these data. For example:**

```{r}
library(caret)
knnModel <- train(x = trainingData$x, y = trainingData$y, method = "knn", preProc = c("center", "scale"), tuneLength = 10)

knnModel

knnPred <- predict(knnModel, newdata = testData$x)

## The function 'postResample' can be used to get the test set
## perforamnce values

postResample(pred = knnPred, obs = testData$y)
```

**Which models appear to give the best performance? Does MARS select the informative predictors (those named X1–X5)?** 

```{r}
knn <- train(x = trainingData$x, 
             y = trainingData$y, 
             method = "knn", 
             preProc = c("center", "scale"), 
             tuneLength = 10)
knn

```
```{r}
svmR <- train(x=trainingData$x, y=trainingData$y, 
                  method="svmRadial", 
                  preProcess=c("center", "scale"), 
                  tuneLength=20)
svmR
```
```{r}
mars_model <- train(x=trainingData$x, 
                    y=trainingData$y, 
                    method="earth",
                    preProcess = c("center", "scale"))

mars_predict <- predict(mars_model, newdata=testData$x)
mars_results <- postResample(pred=mars_predict, obs=testData$y)
mars_results

```

The RMSE Was the best in the MARS model. 

```{r}
varImp(mars_model, 10)
```

Within the MARS model, ONLY the X1-X5 were important, each of the others were effectively discarded. 

# 7.5

**Exercise 6.3 describes data for a chemical manufacturing process. Use the same data imputation, data splitting, and pre-processing steps as before and train several nonlinear regression models.**

## (a) 

**Which nonlinear regression model gives the optimal re-sampling and test set performance?**

```{r}
knn_CMP <- train(
  Yield ~ ., data = CMP.train, method = "knn",
  center = TRUE,
  scale = TRUE,
  trControl = trainControl("cv", number = 10),
  tuneLength = 25
)

knn_CMP
```
```{r}
knn_predictions <- predict(knn_CMP, CMP.test)

postResample(pred = knn_predictions, obs = CMP.test$Yield)
```

```{r}
svmRCMP <- train(
            Yield ~ ., data = CMP.train, method = "svmRadial",
            center = TRUE,
            scale = TRUE,
            trControl = trainControl(method = "cv"),
            tuneLength = 25
          )


svmRCMP


```
```{r}
svm_predictions <- predict(svmRCMP, CMP.test)
postResample(pred = svm_predictions, obs = CMP.test$Yield)

```

```{r eval=FALSE}
nnetGrid <- expand.grid(.decay=c(0, 0.01, 0.1),
                        .size=c(1, 5, 10),
                        .bag=FALSE)
nnetCMP <- train(Yield ~ ., 
                   data = CMP.train,
                   method = "avNNet",
                   tuneGrid = nnetGrid,
                   preProc = c("center", "scale"),
                   trace=FALSE,
                   linout=TRUE,
                   maxit=500)

nnetCMP


```
```{r}
#nnet_predictions <- predict(nnetCMP, CMP.test)
#postResample(pred = nnet_predictions, obs = CMP.test$Yield)

```


## (b) 

**Which predictors are most important in the optimal nonlinear regression model? Do either the biological or process variables dominate the list? How do the top ten important predictors compare to the top ten predictors from the optimal linear model?**

The importance was identical for these models and the linear model. 

```{r}
varImp(knn_CMP)
varImp(svmRCMP)
#varIMP(nnetCMP)

```

## (c) 

**Explore the relationships between the top predictors and the response for the predictors that are unique to the optimal nonlinear regression model. Do these plots reveal intuition about the biological or process predictors and their relationship with yield?**

# 8.1

**Recreate the simulated data from Exercise 7.2:**
```{r}
library(mlbench)
set.seed(200)
simulated <- mlbench.friedman1(200, sd = 1)
simulated <- cbind(simulated$x, simulated$y)
simulated <- as.data.frame(simulated)
colnames(simulated)[ncol(simulated)] <- "y"
```


## (a) 
**Fit a random forest model to all of the predictors, then estimate the variable importance scores:**
```{r}
library(randomForest)
library(caret)
model1 <- randomForest(y ~ ., data = simulated, importance = TRUE, ntree = 1000)
rfImp1 <- varImp(model1, scale = FALSE)
rfImp1
```

**Did the random forest model significantly use the uninformative predictors (V6 – V10)?**

No, the importance drops down significantly after V5. 

## (b)
**Now add an additional predictor that is highly correlated with one of the informative predictors. For example:**

```{r}
simulated$duplicate1 <- simulated$V1 + rnorm(200) * .1
cor(simulated$duplicate1, simulated$V1)
```

**Fit another random forest model to these data. Did the importance score for V1 change? What happens when you add another predictor that is also highly correlated with V1?**

```{r}
model2 <- randomForest(y ~ ., data = simulated, importance = TRUE, ntree = 1000)
rfImp2 <- varImp(model2, scale = FALSE)
rfImp2
```
We see that V1 becomes relatively less important - third on our list. While Duplicate 1 is now ranked 4th. 

## (c) 

**Use the cforest function in the party package to fit a random forest model using conditional inference trees. The party package function varimp can calculate predictor importance. The conditional argument of that function toggles between the traditional importance measure and the modified  version described in Strobl et al. (2007). Do these importances show the same pattern as the traditional random forest model?**

```{r}
library(party)

cforest1 <- party::cforest(y ~ ., data = simulated[, c(1:11)],
                    controls = cforest_control(ntree = 1000))

cforest2 <- party::cforest(y ~ ., data =  simulated,
                    controls = cforest_control(ntree = 1000))

varImp(cforest1, conditional = FALSE)
varImp(cforest2, conditional = FALSE)
#varImp(cforest1, conditional = TRUE)
#varImp(cforest2, conditional = TRUE)


```

When changing Conditional Interference to "on" we see a huge drop in importance in V1 and Duplicate1 in the model which uses each of these variables. V5 also seems to suffer. 

## (d) 

**Repeat this process with different tree models, such as boosted trees and Cubist. Does the same pattern occur?**

```{r}
library(Cubist)

cubist1 <- cubist(y = simulated$y, x = simulated[, c(1:10)],,committees=10)
cubist2 <- cubist(y = simulated$y, x = simulated[, c(1:10,12)],committees=10)

```


```{r}
varImp(cubist1, conditional = FALSE)
varImp(cubist2, conditional = FALSE)
varImp(cubist1, conditional = TRUE)
varImp(cubist2, conditional = TRUE)

```

In the Cubist model, the conditional argument was less impactful. 

# 8.2

**Use a simulation to show tree bias with different granularities**


# 8.3

**In stochastic gradient boosting the bagging fraction and learning rate will govern the construction of the trees as they are guided by the gradient. Although the optimal values of these parameters should be obtained through the tuning process, it is helpful to understand how the magnitudes of these parameters affect magnitudes of variable importance. Figure 8.24 provides the variable importance plots for boosting using two extreme values for the bagging fraction (0.1 and 0.9) and the learning rate (0.1 and 0.9) for the solubility data. The left-hand plot has both parameters set to 0.1, and the right-hand plot has both set to 0.9:**

## (a) 

**Why does the model on the right focus its importance on just the first few of predictors, whereas the model on the left spreads importance across more predictors?**

This is related to the bagging fraction and learning rate. As seen in the text, the bagging fraction is essentially the portion of the training sample used by the model. The learning rate is essentially the rate at which the model selects and drop predictors. The model on the right has a higher bagging fraction and learning rate, thus it is focused on a smaller quantity of predictors in an attempt to quickly find the most "optimal" model (though the peak value may be compromised in favor of speed). 

## (b) 

**Which model do you think would be more predictive of other samples?**

The model on the left. 

## (c) 

**How would increasing interaction depth affect the slope of predictor importance for either model in Fig. 8.24?**

As depth is increased, we would expect to see predictive power spread over more variables. The model on the right would likely use more of its available variables, while the predicative power of the model on the left would too increase, as its variables would be reexamined in different combinations.  

# 8.7

**Refer to Exercises 6.3 and 7.5 which describe a chemical manufacturing process. Use the same data imputation, data splitting, and pre-processing steps as before and train several tree-based models:**

## (a) 

**Which tree-based regression model gives the optimal resampling and test set performance?**

```{r}
rpGrid <- expand.grid(maxdepth= seq(1,10,by=1))
rp <- train(x =  CMP.train[, -YieldX], y = CMP.train$Yield,
                       method = "rpart2",
                       metric = "Rsquared",
                       tuneGrid = rpGrid)
      
rfGrid <- expand.grid(mtry=seq(2,38,by=3))

rf <- train(x =  CMP.train[, -YieldX], y = CMP.train$Yield,
                    method = "rf",
                    tuneGrid = rfGrid,
                    metric = "Rsquared",
                    importance = TRUE)

gbmGrid <- expand.grid(interaction.depth=seq(1,6,by=1),
                       n.trees=c(25,50,100,200),
                       shrinkage=c(0.01,0.05,0.1,0.2),
                       n.minobsinnode=5)

gbm <- train(x =  CMP.train[, -YieldX], y = CMP.train$Yield,
                     method = "gbm",
                     metric = "Rsquared",
                     verbose = FALSE,
                     tuneGrid = gbmGrid)
```

```{r}
cubistCMPGrid <- expand.grid(committees = c(1, 5, 10, 20, 50, 100), 
                          neighbors = c(0, 1, 3, 5, 7))

cubistCMP <- train(x =  CMP.train[, -YieldX], y = CMP.train$Yield,
                        method = "cubist", 
                        verbose = FALSE,
                        metric = "Rsquared",
                        tuneGrid = cubistCMPGrid)
```

```{r}

max(rp$results$RMSE)

max(rf$results$RMSE)

max(gbm$results$RMSE)

max(cubistCMP$results$RMSE)

```

The Random forest Model performs the best of the above. 

## (b) 

**Which predictors are most important in the optimal tree-based regression model? Do either the biological or process variables dominate the list? How do the top 10 important predictors compare to the top 10 predictors from the optimal linear and nonlinear models?**

There are some similarities. ManufacturingProcess32 is the most important in all models.  ManufacturingProcess06 seems to be fairly important in  each of these as well, though this is not as consistent. Otherwise, in general Manufacturing is more important than Biological Materials, however, the Random Forest model puts more importance on Biological Methods than the other models. 

```{r}
varImp(rf)
varImp(modelx)
varImp(knn_CMP)



```

## (c) 

**Plot the optimal single tree with the distribution of yield in the terminal nodes. Does this view of the data provide additional knowledge about the biological or process predictors and their relationship with yield?**

```{r}

```